An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets

### 2021-10-14 ju-hofmann/ju-hofmann.tex
proof read

In this paper we present an open-source implementation of the algorithm
of the Congruent Matching Profile Segments (CMPS) method. \citet{cmps}
developed the CMPS method for ``objective comparison of striated tool
marks'' and demonstrated its use in some examples of comparing bullet
signature correlations. Although \citet{cmps} conceptually described the
CMPS algorithm in their paper, the authors did not release an
implementation of their method. Thus, our effort here is to introduce
the CMPS method to the \texttt{R} \citep{R} community and provide an
open-source, publicly available implementation of the algorithm to use,
review, and improve. Our implementation is made available as part of the
\texttt{R} package \pkg{cmpsR} on GitHub at
\url{https://github.com/willju-wangqian/cmpsR}.

According to the Uniform Crime Reporting Program of FBI (Federal Bureau
of Investigation), ``More than 73 percent (73.7) of the homicides for
which the FBI received weapons data in 2019 involved the use of
firearms'' \citep{fbiucr}, and firearm examination could play an
important role in all these cases. \textbackslash hh\{XXX the new data
for 2020 came out on Sep 27 - homicides are up by close to 30\%. We
might want to mention that.\} An important task of firearm examination
is to answer the question of whether two pieces of evidence come from
the same source or whether a piece of evidence matches a sample obtained
from a specific firearm. Here, in particular, we are interested to
determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of
striation marks acquired during the firing process as bullets are
propelled through the barrel. Current state of art sees firearms
examiners make an assessment of similarity based on a visual comparison,
generally, using a comparison microscope \citep{afte}. This practice has
been criticized for its lack of objectivity and the associated problem
in calculating an error rate \citep{pcast}.

A report published by \citet{nrc} states that ``{[}m{]}uch forensic
evidence-including, for example, bite marks and firearm and toolmark
identification---is introduced in criminal trials without any meaningful
scientific validation, determination of error rates, or reliability
testing to explain the limits of the discipline.'' To overcome those
criticisms and concerns, researchers have been making an effort to build
databases and develop frameworks and algorithms that bring an objective
and quantitative assessment into the field
\citetext{\citealp[\citet{brundage}, \citet{hamby}, \citet{Hamby:2019},
\citet{song2005}, \citet{ChumbleyL_Scott2010VoTM}, \citet{aoas},
\citet{pmid30444940}]{nistdb}; \citealp{cmps}}. Note that this is not an
exhaustive list of these efforts. The notion of bullet signatures is one
of the results of these efforts. Bullet signatures can be extracted from
bullet land engraved areas (LEAs), typically one for each land engraved
area, and how those bullet signatures are extracted from scans will be
discussed in the Background section. Since bullet signatures can
numerically describe the striation marks on the bullet and can be
understood by algorithms, they played an important role in bringing
objectivity into the field. In the following sections we will: review
the background of bullet signature comparisons, discuss how we followed
the idea described by \citet{cmps} for the implementation of the CMPS
algorithm, propose new metrics based on the CMPS score and a principled
evaluation framework for algorithmic results comparison, and present
results of applying our implementation to real data.

@@

The datasets we worked with come from the James Hamby Consecutively
Rifled Ruger Barrel Study \citep{brundage, hamby, Hamby:2019}, in
particular, Hamby set 252 and Hamby set 44. For each Hamby set, a total
of 35 bullets is fired from (the same) ten consecutively manufactured
Ruger P-85 pistol barrels. Two bullets are fired from each barrel,
making up a set of 20 reference bullets. An additional 15 bullets are
fired from these ten barrels in a fashion unknown to the study
participant. The aim of the Hamby Study was to have firearms examiners
identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves
and lands as shown in \autoref{fig:bullet}. During the firing process,
grooves and lands are engraved on a bullet. Firearms examiners use
striation marks on land engraved areas (LEAs) for their visual
comparison. For algorithmic purposes, 3D topographical images of land
engraved areas were obtained and stored in x3p format (XML 3-D Surface
Profile). The x3p format provides a standard way of exchanging 2D and 3D
profile data. It conforms to the ISO5436-2 standard
{[}\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}{]} adopted by the
OpenFMC (Open Forensic Metrology Consortium), a group of firearm
forensics researchers who contributes to the establishment of best
practices of using metrology in forensic science. Hamby set 252 were
scanned using a NanoFocus lens at 20x magnification with the scan
resolution being 1.5625 \textmu m \(\times\) 1.5625 \textmu m per pixel.
Hamby set 44 was scanned at the Roy J Carver High-Resolution microscopy
lab at Iowa State. These scans were acquired with a Sensofar Confocal
Light Microscope at 20x magnification for a nominal resolution of 0.645
\textmu m \(\times\) 0.645 \textmu m per pixel. Both Hamby set 252 and
Hamby set 44 are publicly available on the NIST Ballistics Database
Project \citep{nistdb}.

@@

The automated framework used in this paper was proposed by \citet{aoas}.
In order to obtain bullet signatures from the x3p files of bullet land
engravings, we use the \texttt{R} packages \pkg{x3ptools}
\citep{x3ptools} and \pkg{bulletxtrctr} \citep{bulletxtrctr}.
\pkg{x3ptools} is a package to read, write, and generally, process x3p
files. The \pkg{bulletxtrctr} package implements a pipeline for
extracting and comparing signatures from scans of land-engraved areas.

@@

The white horizontal line in \autoref{fig:process}(a) indicates which
cross-section was identified by the algorithm to represent the LEA;
\autoref{fig:process}(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the
vertical blue lines in \autoref{fig:process}(c) and
\autoref{fig:process}(d)). A non-parametric LOESS smooth \citep{loess}
is fitted to capture the bullet curvature (\autoref{fig:process}(e))
and, finally, the \textbf{bullet signature} (\autoref{fig:process}(f))
is obtained as residuals of the cross-section and the fitted smooth.
Note, that in \citet{cmps} bullet signatures are referred to as bullet
profiles. However, to avoid confusion, we distinguish the notion of
bullet signatures from bullet profiles. Bullet profiles are shown in
panels (b), (c), and (d) of \autoref{fig:process}, while
\autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a
correct down-stream analysis of the signatures. To allow for a
human-in-the middle inspection and intervention we have provided an
interactive web-application, implemented as \texttt{R} Shiny App
\citep{shiny}, \pkg{bulletinspectR} to identify and correct those
errors. An example of the extraction process with corresponding code and
parameter settings can be found on Github at
\url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}.
Note that the process of extracting signatures might be different from
the one used in \citet{cmps} because no code or parameter settings are
made available publicly.

@@

\caption[A framework of obtaining a bullet signature]{A framework of obtaining a bullet signature. (a) front view of a scanned land engraved area (LEA). The selected crosscut location is indicated by the white line. (b) view of the cross-section of the land engraved area at the white line in (a). (c) the crosscut data plotted in 2D; blue vertical lines indicate the position of left and right grooves. (d) the crosscut data after chopping the left and right grooves. (e) the fitted curvature using LOESS. (f) after removing the curvature from the crosscut data, the bullet signature is obtained}\label{fig:process}
\end{figure}
\end{Schunk}

@@

Most algorithms for comparing striation marks are based on the digitized
signatures and produce a similarity score
\citep[\citet{ChumbleyL_Scott2010VoTM}, \citet{aoas},
\citet{pmid30444940}]{song2005}. The congruent matching profile segments
(CMPS) algorithm, developed by \citet{cmps} for ``objective comparison
of striated tool marks'', is one such algorithm. The algorithm's main
idea is to take a set of consecutive and non-overlapping basis segments
from the comparison and for each segment find the ``best'' registration
position on the reference (the other bullet signature) with respect to
their cross-correlation values. From a comparison of these registration
positions, a \textbf{congruent registration position} is identified, and
the number of basis segments taking the congruent registration position
is the CMPS score. Note that researchers in \citet{cmps} refers to the
origin of basis segments as the reference, but in this paper we refer to
it as the comparison. High CMPS scores are achieved between more similar
signatures and are therefore indicative of a same-source pair. Low
scores between pairs of signatures are attributed to different source
pairs. However, a specific threshold of the CMPS score to distinguish
between same-source and different-source comparisons is not provided in
\citet{cmps}, instead the threshold depends on the underlying structure
of the data and the choice of parameters.
\hh{XXX  In a legal setting this variability is problematic because it allows for situations in which experts could choose parameters based on a wished-for outcome. }
Further research is needed to understand how to determine optimal
threshold settings. The CMPS algorithm can assist firearm examiners with
drawing a conclusion about the source of a comparison pair.
Unfortunately, \citet{cmps} did not release any code or specific
parameter settings for their implementation of the CMPS algorithm. Thus,
in this paper we present an open-source implementation of the CMPS
algorithm in the R package \pkg{cmpsR} available from Github
\url{https://github.com/willju-wangqian/cmpsR}. This publicly available
implementation calculates the CMPS score of a comparison using the
following code:

@@

The section ``Implementation'' will go through the algorithm and show
how to use the \pkg{cmpsR} package. A further example that illustrates
the main points is also included. The section ``Results'' presents the
results of evaluating the \pkg{cmpsR} package using Hamby set 252 and
Hamby set 44. The results from Hamby set 252 are used to verify that our
implementation is, at least qualitatively, comparable to the algorithm
described in \citet{cmps}. Results from Hamby 44 show the need for a
further investigation of the parameter choices even in the case of
bullets fired from the same barrels.
\wj{We suggest new CMPS metrics that summarize land-level CMPS scores and a sum of squares ratio that can be used to evaluate algorithmic results.}
The last section covers some final discussion and conclusions.

@@

\caption[CMPS scores of all 36 pairwise bullet signature comparisons for two bullets]{CMPS scores of all 36 pairwise bullet signature comparisons for two bullets. Land engraving pairs generated by the same land (KM comparisons) are highlighted}\label{fig:tiles}
\end{figure}
\end{Schunk}

@@

\address{%
Wangqian W. Ju\\
Department of Statistics\\%
Center for Statistics and Applications in Forensic Evidence\\ Iowa State
University\\ 2438 Osborn Dr\\ Ames, IA 50011\\
%
\url{https://github.com/willju-wangqian}%
\\\textit{ORCiD: \href{https://orcid.org/0000-0002-9977-377X}{0000-0002-9977-377X}}%
\\\href{mailto:wju@iastate.edu}{\nolinkurl{wju@iastate.edu}}
}

\address{%
Heike Hofmann\\
Department of Statistics\\%
Center for Statistics and Applications in Forensic Evidence\\ Iowa State
University\\ 2438 Osborn Dr\\ Ames, IA 50011\\
%
\url{https://github.com/heike}%
\\\textit{ORCiD: \href{https://orcid.org/0000-0002-9079-593X}{0000-0002-9079-593X}}%
\\\href{mailto:hofmann@iastate.edu}{\nolinkurl{hofmann@iastate.edu}}
}

>>>>

In this paper we present an open-source implementation of the algorithm
of the Congruent Matching Profile Segments (CMPS) method. \citet{cmps}
developed the CMPS method for ``objective comparison of striated tool
marks'' and demonstrated its use in some examples of comparing bullet
signature correlations. Although \citet{cmps} conceptually describe the
CMPS algorithm in their paper, the authors did not release an
implementation of their method. Thus, our effort here is to introduce
the CMPS method to the \texttt{R} \citep{R} community and provide an
open-source, publicly available implementation of the algorithm to use,
review, and improve. Our implementation is made available as part of the
\texttt{R} package \pkg{cmpsR} on GitHub at
\url{https://github.com/willju-wangqian/cmpsR}.

According to the Uniform Crime Reporting Program of FBI (Federal Bureau
of Investigation), ``More than 73 percent (73.7) of the homicides for
which the FBI received weapons data in 2019 involved the use of
firearms'' \citep{fbiucr}, and firearm examination could play an
important role in all these cases. \textbackslash hh\{XXX the new data
for 2020 came out on Sep 27 - homicides are up by close to 30\%. We
might want to mention that.\} An important task of firearm examination
is to answer the question of whether two pieces of evidence come from
the same source or whether a piece of evidence matches a sample obtained
from a specific firearm. Here, in particular, we are interested to
determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of
striation marks acquired during the firing process as bullets are
propelled through the barrel. Current state of art sees firearms
examiners make an assessment of similarity based on a visual comparison,
generally, using a comparison microscope \citep{afte}. This practice has
been criticized for its lack of objectivity and the associated problem
of determining valid error rates \citep{pcast}.

A report published by the \citet{nrc} states that ``{[}m{]}uch forensic
evidence-including, for example, bite marks and firearm and toolmark
identification---is introduced in criminal trials without any meaningful
scientific validation, determination of error rates, or reliability
testing to explain the limits of the discipline.'' To overcome those
criticisms and concerns, researchers have been making an effort to build
databases and develop frameworks and algorithms that bring an objective
and quantitative assessment into the field
\hh{XXX could you summarize the main contribution of each of these papers in at most a sentence?}
\citetext{\citealp[\citet{brundage}, \citet{hamby}, \citet{Hamby:2019},
\citet{song2005}, \citet{ChumbleyL_Scott2010VoTM}, \citet{aoas},
\citet{pmid30444940}]{nistdb}; \citealp{cmps}}. Note that this is not an
exhaustive list of these efforts. The notion of bullet signatures is one
of the results of these efforts. Bullet signatures can be extracted from
bullet land engraved areas (LEAs), typically one for each land engraved
area, and how those bullet signatures are extracted from scans will be
discussed in the Background section. Since bullet signatures can
numerically describe the striation marks on the bullet and can be
understood by algorithms, they played an important role in bringing
objectivity into the field. In the following sections we will: review
the background of bullet signature comparisons, discuss how we followed
the idea described by \citet{cmps} for the implementation of the CMPS
algorithm, propose new metrics based on the CMPS score and a principled
evaluation framework for algorithmic results comparison, and present
results of applying our implementation to real data.

@@

The datasets we worked with come from the James Hamby Consecutively
Rifled Ruger Barrel Study \citep{brundage, hamby, Hamby:2019}, in
particular, Hamby set 252 and Hamby set 44. For each Hamby set, a total
of 35 bullets is fired from (the same) ten consecutively manufactured
Ruger P-85 pistol barrels. Two bullets are fired from each barrel,
making up a set of 20 reference bullets. An additional 15 bullets are
fired from these ten barrels in a fashion unknown to the study
participant. The aim of the Hamby Study was to have firearms examiners
identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves
and lands as shown in \autoref{fig:bullet}. During the firing process,
grooves and lands are engraved on a bullet. Firearms examiners use
striation marks on land engraved areas (LEAs) for their visual
comparison. For algorithmic purposes, 3D topographical images of land
engraved areas were obtained and stored in x3p format (XML 3-D Surface
Profile). The x3p format provides a standard way of exchanging 2D and 3D
profile data. It conforms to the ISO5436-2 standard
{[}\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}{]} adopted by the
OpenFMC (Open Forensic Metrology Consortium), a group of firearm
forensics researchers who contributes to the establishment of best
practices of using metrology in forensic science. Hamby set 252 was
scanned using a NanoFocus lens at 20x magnification with the scan
resolution being 1.5625 \textmu m \(\times\) 1.5625 \textmu m per pixel.
Hamby set 44 was scanned at the Roy J Carver High-Resolution microscopy
lab at Iowa State. These scans were acquired with a Sensofar Confocal
Light Microscope at 20x magnification for a nominal resolution of 0.645
\textmu m \(\times\) 0.645 \textmu m per pixel. Both Hamby set 252 and
Hamby set 44 are publicly available from the NIST Ballistics Database
Project \citep{nistdb}.

@@

The automated framework for extracting signatures from x3p files used in
this paper was proposed by \citet{aoas} and is implemented in the
\texttt{R} packages \pkg{x3ptools} \citep{x3ptools} and
\pkg{bulletxtrctr} \citep{bulletxtrctr}. \pkg{x3ptools} is a package to
read, write, and generally, process x3p files. The \pkg{bulletxtrctr}
package implements a pipeline for extracting and comparing signatures
from scans of land-engraved areas.

@@

The thin white horizontal line in \autoref{fig:process}(a) indicates
which cross-section was identified by the algorithm to represent the
LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional
view. Groove engraved areas are removed from the analysis as indicated
by the vertical blue lines in \autoref{fig:process}(c) and
\autoref{fig:process}(d)). A non-parametric LOESS smooth \citep{loess}
is fitted to capture the bullet curvature (\autoref{fig:process}(e))
and, finally, the \textbf{bullet signature} (\autoref{fig:process}(f))
is obtained as residuals of the cross-section and the fitted smooth.
Note, that in \citet{cmps} bullet signatures are referred to as bullet
profiles. However, to avoid confusion, we distinguish the notion of
bullet signatures from bullet profiles. Bullet profiles are shown in
panels (b), (c), and (d) of \autoref{fig:process}, while
\autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a
correct down-stream analysis of the signatures. To allow for a
human-in-the middle inspection and intervention we have provided an
interactive web-application, implemented as \texttt{R} Shiny App
\citep{shiny}, \pkg{bulletinspectR} to identify and correct those
errors. An example of the extraction process with corresponding code and
parameter settings can be found on Github at
\url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}.
Note that the process of extracting signatures might be different from
the one used in \citet{cmps} because no code or parameter settings are
made available publicly.

@@

\caption[A framework of obtaining a bullet signature]{A framework of obtaining a bullet signature. (a) rendering from the 3d topographic scan of a land engraved area (LEA). The selected crosscut location is indicated by a thin white horizontal line. (b) view of the cross-section of the land engraved area at the white line in (a). (c) the crosscut data plotted in 2D; blue vertical lines indicate the position of left and right grooves. (d) the crosscut data after chopping the left and right grooves. (e) the fitted curvature using LOESS. (f) after removing the curvature from the crosscut data, the bullet signature is obtained}\label{fig:process}
\end{figure}
\end{Schunk}

@@

Most algorithms for comparing striation marks are based on the digitized
signatures and produce a similarity score
\citep[\citet{ChumbleyL_Scott2010VoTM}, \citet{aoas},
\citet{pmid30444940}]{song2005}. The congruent matching profile segments
(CMPS) algorithm, developed by \citet{cmps} for ``objective comparison
of striated tool marks'', is one such algorithm. The algorithm's main
idea is to take a set of consecutive and non-overlapping basis segments
from the comparison and for each segment find the ``best'' registration
position on the reference (the other bullet signature) with respect to
their cross-correlation values. From a comparison of these registration
positions, a \textbf{congruent registration position} is identified, and
the number of basis segments taking the congruent registration position
is the CMPS score. Note that researchers in \citet{cmps} refer to the
origin of basis segments as the reference, but in this paper we refer to
it as the comparison signature. High CMPS scores are achieved between
more similar signatures and are therefore indicative of a same-source
pair. Low scores between pairs of signatures are attributed to different
source pairs. However, a specific threshold of the CMPS score to
distinguish between same-source and different-source comparisons is not
provided in \citet{cmps}, instead the threshold depends on the
underlying structure of the data and the choice of parameters. In a
legal setting this variability is problematic because it allows for
situations in which experts could choose parameters based on whether
they are witnesses for the defense or the prosecution. Further research
is needed to understand how to determine optimal threshold settings. The
CMPS algorithm can assist firearm examiners with drawing a conclusion
about the source of a comparison pair. Thus, in this paper we present an
open-source implementation of the CMPS algorithm in the R package
\pkg{cmpsR} available from Github
\url{https://github.com/willju-wangqian/cmpsR}. This publicly available
implementation calculates the CMPS score of a comparison using the
following code:

@@

The section ``Implementation'' will go through the algorithm and show
how to use the \pkg{cmpsR} package. A further example that illustrates
the main points is also included. In the section on 'Evaluation metrics"
we introduce new CMPS metrics that summarize land-level CMPS scores and
a sum of squares ratio that can be used to evaluate algorithmic results.
The section ``Results'' presents the results of evaluating the
\pkg{cmpsR} package using Hamby set 252 and Hamby set 44. The results
from Hamby set 252 are used to verify that our implementation is, at
least qualitatively, comparable to the algorithm described in
\citet{cmps}. Results from Hamby 44 show the need for a further
investigation of the parameter choices even in the case of bullets fired
from the same barrels. The last section covers some final discussion and
conclusions.

@@

\caption{CMPS scores of all 36 pairwise bullet signature comparisons for two bullets. Land engraving pairs generated by the same land (KM comparisons) are highlighted. \hh{Note that in this example the axis along Bullet 2 starts with Land 2. This corresponds to Phase 2 in equation XXX}}\label{fig:tiles}
\end{figure}
\end{Schunk}

@@

\address{%
Wangqian W. Ju\\
Department of Statistics\\%
Center for Statistics and Applications in Forensic Evidence\\ Iowa State
University\\ 2438 Osborn Dr\\ Ames, IA 50011\\
%
\url{https://github.com/willju-wangqian}\\%
\textit{ORCiD: \href{https://orcid.org/0000-0002-9977-377X}{0000-0002-9977-377X}}\\%
\href{mailto:wju@iastate.edu}{\nolinkurl{wju@iastate.edu}}%
}

\address{%
Heike Hofmann\\
Department of Statistics\\%
Center for Statistics and Applications in Forensic Evidence\\ Iowa State
University\\ 2438 Osborn Dr\\ Ames, IA 50011\\
%
\url{https://github.com/heike}\\%
\textit{ORCiD: \href{https://orcid.org/0000-0002-9079-593X}{0000-0002-9079-593X}}\\%
\href{mailto:hofmann@iastate.edu}{\nolinkurl{hofmann@iastate.edu}}%
}
